# Data structure and preparation {#sec-data}

```{r}
#| include: false
source("_common.R")
```

This chapter runs **live**: it reads the raw spreadsheet and builds the
estimation dataset, exactly the data-preparation workflow demonstrated in class.

## The raw file

The authors provided `harrison.xls`. The relevant sheet, `trt 10&11`, is a
*wide* spreadsheet — one row per respondent — with a two-row header, label rows
separating the two treatments, and summary rows at the bottom. Treatment 10 is
the nonhypothetical group; Treatment 11 is the hypothetical group.

```{r raw-peek}
raw_dims <- dim(readxl::read_excel("harrison.xls", sheet = "trt 10&11",
                                   skip = 3, col_types = "text"))
cat("Raw sheet (after skipping 3 header rows):",
    raw_dims[1], "rows ×", raw_dims[2], "columns\n")
```

**Key variables.**

| Block | Columns | Contents |
|---|---|---|
| Identifiers | `subject`, `session` | Subject ID (1001–1099 = nonhyp; 1100–1199 = hyp) |
| Demographics | `gender`, `age`, `education`, `student`, `income`, … | Used for Table 1 |
| Attitudes | `color`, `tenderness`, `flavor`, … | Survey Likert items (not used in estimation) |
| **Choices** | `sc1`–`sc17` | Chosen alternative per scenario, coded 1–6 |

Choice codes: `1 = Generic (GEN)`, `2 = Guaranteed Tender (GT)`,
`3 = Natural (NAT)`, `4 = USDA Choice (CHO)`, `5 = CAB`, `6 = None (NONE)`.

## The price design (paper appendix)

The 17 × 5 price matrix is the experimental design; it is the same for both
treatments and is the bridge between a chosen *code* and the *prices* the
respondent faced.

```{r price-matrix}
price_matrix <- tibble::tribble(
  ~scenario, ~GEN,  ~GT,   ~NAT,  ~CHO,  ~CAB,
  1, 3.38, 7.88, 6.75, 6.75, 6.75,   2, 4.50, 9.00, 6.75, 5.63, 9.00,
  3, 6.75, 9.00, 5.63, 9.00, 6.75,   4, 5.63, 9.00, 9.00, 6.75, 5.63,
  5, 6.75, 6.75, 6.75, 7.88, 7.88,   6, 4.50, 7.88, 7.88, 9.00, 9.00,
  7, 4.50, 6.75, 5.63, 6.75, 7.88,   8, 5.63, 5.63, 5.63, 7.88, 9.00,
  9, 5.63, 7.88, 5.63, 7.88, 9.00,  10, 5.63, 6.75, 7.88, 5.63, 9.00,
 11, 6.75, 7.88, 9.00, 5.63, 7.88,  12, 4.50, 5.63, 5.63, 7.88, 7.88,
 13, 3.38, 9.00, 7.88, 7.88, 9.00,  14, 3.38, 5.63, 5.63, 5.63, 5.63,
 15, 3.38, 6.75, 9.00, 9.00, 9.00,  16, 6.75, 5.63, 7.88, 6.75, 6.75,
 17, 5.63, 5.63, 5.63, 5.63, 5.63
)
kable(price_matrix, caption = "Price design: 17 scenarios × 5 steaks (USD).")
```

## Cleaning: from spreadsheet to respondents

I keep only valid subject-ID rows (1001–1199), which automatically drops the
treatment-label rows and the trailing summary rows. **One subject (ID 1060) has
all 17 scenarios missing and is excluded** — this reconciles the 68 rows in the
file with the 67 nonhypothetical subjects the paper reports.

```{r clean-wide}
demo_names <- c(
  "subject","gender","age","household","child","education","income","student",
  "occupation","race","gbeef","steak_buy","other_beef","gbeef_freq","steak_freq",
  "color","brand","quality_gd","ex_fat","in_fat","price_imp","safety","juiciness",
  "flavor","tenderness","consistency","doneness","production","quality_gd2",
  "slaughter","food_safe","quality_pref","cooked","chance1","chance2","chance3",
  "chance4","chance5")
sc_names <- paste0("sc", 1:17)

raw <- read_excel("harrison.xls", sheet = "trt 10&11",
                  col_names = c(demo_names, sc_names, "blank", "session"),
                  skip = 3, col_types = "text")

df_wide <- raw |>
  mutate(subject = suppressWarnings(as.numeric(subject))) |>
  filter(!is.na(subject), subject >= 1001, subject <= 1199) |>
  filter(subject != 1060) |>                              # all scenarios missing
  mutate(treatment = factor(if_else(subject < 1100, "nonhyp", "hyp"),
                            levels = c("nonhyp", "hyp"))) |>
  mutate(across(all_of(c(setdiff(demo_names, "subject"), sc_names, "session")),
                as.numeric))

cat("Nonhypothetical subjects:", sum(df_wide$treatment == "nonhyp"), "(paper: 67)\n")
cat("Hypothetical  subjects:",  sum(df_wide$treatment == "hyp"),    "(paper: 37)\n")
```

## Reshape: wide → long → choice format

Two reshapes turn the respondent-level table into the estimation data:

1. **Long** — one row per *respondent × scenario*; merge in the five prices.
2. **Choice (mlogit) format** — one row per *respondent × scenario ×
   alternative*, with a logical `chosen` flag and the alternative's own price.
   **Scenario 17 is excluded from estimation** (it is the all-equal-price binding
   scenario), leaving scenarios 1–16.

```{r reshape}
df_long <- df_wide |>
  select(subject, treatment, session, all_of(sc_names)) |>
  pivot_longer(all_of(sc_names), names_to = "sc_var", values_to = "choice_code") |>
  mutate(scenario = as.integer(sub("sc", "", sc_var)),
         choice   = factor(choice_code, levels = 1:6, labels = alt_labels)) |>
  select(-sc_var, -choice_code) |>
  left_join(price_matrix, by = "scenario") |>
  arrange(subject, scenario)

df_mlogit <- df_long |>
  filter(scenario <= 16) |>                               # drop binding scenario
  crossing(alt = factor(alt_labels, levels = alt_labels)) |>
  mutate(
    chosen    = as.character(choice) == as.character(alt),
    price_alt = case_when(
      alt == "GEN" ~ GEN, alt == "GT"  ~ GT,  alt == "NAT" ~ NAT,
      alt == "CHO" ~ CHO, alt == "CAB" ~ CAB, alt == "NONE" ~ NA_real_),
    chid = paste(subject, scenario, sep = "_")) |>
  select(chid, subject, treatment, session, scenario, alt, chosen, price_alt) |>
  arrange(subject, scenario, alt)

cat("Nonhyp choice situations:",
    sum(df_mlogit$treatment == "nonhyp" & df_mlogit$alt == "GEN"), "(paper: 1072 = 67×16)\n")
cat("Hyp    choice situations:",
    sum(df_mlogit$treatment == "hyp"    & df_mlogit$alt == "GEN"), "(paper:  592 = 37×16)\n")
```

```{r show-choice-format}
df_mlogit |> filter(chid == "1001_1") |>
  select(chid, alt, chosen, price_alt) |>
  kable(caption = "One choice situation in long format (subject 1001, scenario 1).")
```

::: {.callout-note}
## Data-prep lessons
* The "68 vs 67" gap is the first thing a replicator hits. Rather than guess, I
  searched for the structural reason — subject 1060 is empty — which is exactly
  the kind of provenance note the original LIMDEP file does not carry.
* `NONE` has **no price**; encoding its `price_alt` as `NA`/0 (and never giving
  it an own-price coefficient) is what makes it the natural base alternative.
:::
