# Descriptive replication: Tables 1 and 2 {#sec-descriptives}

```{r}
#| include: false
source("_common.R")
```

## Table 1 — demographics by treatment

The paper's Table 1 shows that randomisation worked: means are statistically
indistinguishable across treatments. I reproduce the means, SDs, and t-tests.

```{r table1}
mean_sd <- function(data, var) {
  data |> group_by(treatment) |>
    summarise(mean = round(mean(.data[[var]], na.rm = TRUE), 2),
              sd   = round(sd(.data[[var]],   na.rm = TRUE), 2), .groups = "drop") |>
    pivot_wider(names_from = treatment, values_from = c(mean, sd)) |>
    mutate(variable = var)
}
t1_vars <- c("gender", "age", "education", "student", "income")
table1 <- bind_rows(lapply(t1_vars, mean_sd, data = df_wide)) |>
  rowwise() |>
  mutate(t_p = round(t.test(df_wide[[variable]][df_wide$treatment == "nonhyp"],
                            df_wide[[variable]][df_wide$treatment == "hyp"])$p.value, 3)) |>
  ungroup() |>
  select(variable, mean_nonhyp, sd_nonhyp, mean_hyp, sd_hyp, t_p)

kable(table1,
      col.names = c("Variable", "NH mean", "NH (SD)", "HY mean", "HY (SD)", "t-test p"),
      caption = "Table 1. Demographics by treatment (NH = nonhypothetical, HY = hypothetical).")
```

::: {.callout-tip}
## Result
Every t-test p-value exceeds 0.05 — as in the paper, we cannot reject equality
of demographic means across treatments. Randomisation is validated, so any
behavioural difference later is attributable to the hypothetical/real
manipulation, not to sample composition.
:::

## Table 2 — choice frequencies and the "None" gap

```{r table2}
table2 <- df_long |>
  count(treatment, scenario, choice) |>
  group_by(treatment, scenario) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |> select(-n) |>
  pivot_wider(names_from = choice, values_from = pct, values_fill = 0) |>
  select(treatment, scenario, all_of(alt_labels)) |>
  arrange(treatment, scenario)

kable(table2 |> filter(treatment == "nonhyp") |> select(-treatment),
      col.names = c("Scenario","Generic","GT","Natural","USDA Choice","CAB","None"),
      caption = "Table 2a. % choosing each alternative — Nonhypothetical.")
kable(table2 |> filter(treatment == "hyp") |> select(-treatment),
      col.names = c("Scenario","Generic","GT","Natural","USDA Choice","CAB","None"),
      caption = "Table 2b. % choosing each alternative — Hypothetical.")
```

```{r none-gap}
none_gap <- table2 |> filter(scenario <= 16) |>
  select(treatment, scenario, NONE) |>
  pivot_wider(names_from = treatment, values_from = NONE) |>
  mutate(gap = nonhyp - hyp)
cat(sprintf("Smallest None gap: scenario %d (%.1f pp)\n",
            none_gap$scenario[which.min(abs(none_gap$gap))], min(abs(none_gap$gap))))
cat(sprintf("Largest  None gap: scenario %d (%.1f pp)\n",
            none_gap$scenario[which.max(none_gap$gap)], max(none_gap$gap)))
cat("Paper: smallest = scenario 4 (8.31 pp), largest = scenario 7 (32.84 pp)\n")
```

::: {.callout-tip}
## Result — the core stylised fact
Nonhypothetical subjects pick **"None"** much more often than hypothetical
subjects (a positive gap in every scenario). This is hypothetical bias in its
rawest form: when payment is real, more people walk away. The gap reproduces the
paper's reported extremes at scenarios 4 and 7 almost exactly.
:::
