# Descriptive replication: Tables 1 and 2 {#sec-descriptives}

```{r}
#| include: false
source("_common.R")
```

## Table 1 — demographics by treatment

Table 1 in the paper is a randomisation check: it shows that the two treatment
groups look alike on observables. I reproduce the means, standard deviations, and
the t-tests for equality of means.

```{r table1}
mean_sd <- function(data, var) {
  data |> group_by(treatment) |>
    summarise(mean = round(mean(.data[[var]], na.rm = TRUE), 2),
              sd   = round(sd(.data[[var]],   na.rm = TRUE), 2), .groups = "drop") |>
    pivot_wider(names_from = treatment, values_from = c(mean, sd)) |>
    mutate(variable = var)
}
t1_vars <- c("gender", "age", "education", "student", "income")
table1 <- bind_rows(lapply(t1_vars, mean_sd, data = df_wide)) |>
  rowwise() |>
  mutate(t_p = round(t.test(df_wide[[variable]][df_wide$treatment == "nonhyp"],
                            df_wide[[variable]][df_wide$treatment == "hyp"])$p.value, 3)) |>
  ungroup() |>
  select(variable, mean_nonhyp, sd_nonhyp, mean_hyp, sd_hyp, t_p)

kable(table1,
      col.names = c("Variable", "NH mean", "NH (SD)", "HY mean", "HY (SD)", "t-test p"),
      caption = "Table 1. Demographics by treatment (NH = nonhypothetical, HY = hypothetical).")
```

::: {.callout-tip}
## Result
None of the t-tests is significant at 5%, so I cannot reject equality of the
demographic means across treatments, just as in the paper. Because the groups are
balanced, any difference in behaviour later on can be read as coming from the
hypothetical-versus-real manipulation rather than from who happened to be in each
group.
:::

## Table 2 — choice frequencies and the "None" gap

```{r table2}
table2 <- df_long |>
  count(treatment, scenario, choice) |>
  group_by(treatment, scenario) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |> select(-n) |>
  pivot_wider(names_from = choice, values_from = pct, values_fill = 0) |>
  select(treatment, scenario, all_of(alt_labels)) |>
  arrange(treatment, scenario)

kable(table2 |> filter(treatment == "nonhyp") |> select(-treatment),
      col.names = c("Scenario","Generic","GT","Natural","USDA Choice","CAB","None"),
      caption = "Table 2a. % choosing each alternative — Nonhypothetical.")
kable(table2 |> filter(treatment == "hyp") |> select(-treatment),
      col.names = c("Scenario","Generic","GT","Natural","USDA Choice","CAB","None"),
      caption = "Table 2b. % choosing each alternative — Hypothetical.")
```

```{r none-gap}
none_gap <- table2 |> filter(scenario <= 16) |>
  select(treatment, scenario, NONE) |>
  pivot_wider(names_from = treatment, values_from = NONE) |>
  mutate(gap = nonhyp - hyp)
cat(sprintf("Smallest None gap: scenario %d (%.1f pp)\n",
            none_gap$scenario[which.min(abs(none_gap$gap))], min(abs(none_gap$gap))))
cat(sprintf("Largest  None gap: scenario %d (%.1f pp)\n",
            none_gap$scenario[which.max(none_gap$gap)], max(none_gap$gap)))
cat("Paper: smallest = scenario 4 (8.31 pp), largest = scenario 7 (32.84 pp)\n")
```

::: {.callout-tip}
## Result: the basic pattern
In every scenario, the nonhypothetical group picks "None" more often than the
hypothetical group. When the payment is real, more people simply walk away. The
size of that gap lines up with the paper, which is largest at scenario 7 and
smallest at scenario 4.
:::
