# Multinomial logit and the scale test {#sec-mnl}

```{r}
#| include: false
source("_common.R")
```

::: {.callout-note}
## How the model chapters work
From here on the estimation code is shown but not re-run when the book builds. It
is the code in `R/03`–`R/06`. The fitted objects are loaded from `output/`, and
the results are recomputed for display. Each chapter follows the same order:
specification, estimation code, results, then my comments.
:::

## Specification

The utility of alternative $j$ in choice situation $i$ is

$$
V_{ij} = \beta_j + \alpha_j P_{ij}, \qquad j \in \{\text{GEN, GT, NAT, CHO, CAB}\},
\quad V_{i,\text{NONE}} \equiv 0 .
$$

So there are five alternative-specific constants $\beta_j$ and five own-price
coefficients $\alpha_j$, ten taste parameters in all. The joint model stacks both
treatments, forces the taste parameters to be equal, and scales the
nonhypothetical utilities by $\mu$:

$$
V^{\text{NH}}_{ij} = \mu\,(\beta_j + \alpha_j P_{ij}), \qquad
V^{\text{H}}_{ij}  =      \beta_j + \alpha_j P_{ij}.
$$

Here $\mu$ is the error scale of the real-payment group relative to the
hypothetical group. The Swait–Louviere LR test compares this joint model with the
two separate models.

## Estimation code (`R/03_table3_mnl.R`)

```{r mnl-code, eval=FALSE}
# Alternative-specific price variables: p_j = price × 1{alt = j}
df0 <- df_mlogit |>
  mutate(price_alt = tidyr::replace_na(price_alt, 0),
         p_GEN = if_else(alt=="GEN", price_alt, 0),
         p_GT  = if_else(alt=="GT",  price_alt, 0),
         p_NAT = if_else(alt=="NAT", price_alt, 0),
         p_CHO = if_else(alt=="CHO", price_alt, 0),
         p_CAB = if_else(alt=="CAB", price_alt, 0))

make_mdata <- function(d) mlogit.data(d, choice="chosen", shape="long",
  alt.var="alt", chid.var="chid", id.var="subject", alt.levels=alts)

f_mnl <- chosen ~ p_GEN + p_GT + p_NAT + p_CHO + p_CAB | 1   # 5 prices + 5 ASCs
mnl_nh  <- mlogit(f_mnl, make_mdata(filter(df0, treatment=="nonhyp")), reflevel="NONE")
mnl_hyp <- mlogit(f_mnl, make_mdata(filter(df0, treatment=="hyp")),    reflevel="NONE")

# Joint model with scale mu: custom log-likelihood maximised by optim(BFGS).
# theta = (beta[1:5], alpha[6:10], log_mu[11]); V_NH scaled by exp(log_mu).
ll_joint_fn <- function(theta, df) {
  beta  <- c(theta[1:5], 0); alpha <- c(theta[6:10], 0); mu <- exp(theta[11])
  k <- match(as.character(df$alt), alts)
  V <- beta[k] + alpha[k]*df$price_alt
  V[df$treatment=="nonhyp"] <- mu * V[df$treatment=="nonhyp"]
  df |> mutate(V=V) |> group_by(chid) |>
    mutate(p = exp(V-max(V))/sum(exp(V-max(V)))) |> ungroup() |>
    summarise(ll = sum(log(pmax(p[chosen], 1e-300)))) |> pull(ll)
}
opt <- optim(theta0, ll_joint_fn, df=df0, method="BFGS",
             control=list(fnscale=-1, maxit=1e4, reltol=1e-12), hessian=TRUE)

# Swait–Louviere LR scale test: chi2 = -2(LL_joint - LL_nh - LL_hyp)
lr_stat <- -2 * (opt$value - logLik(mnl_nh) - logLik(mnl_hyp))
```

There is no off-the-shelf `mlogit` command for the scaled joint model, so I wrote
the log-likelihood by hand and maximised it with `optim`. The `exp(V - max(V))`
inside the probability is just the log-sum-exp trick to keep things numerically
stable.

## Results

```{r mnl-results}
kable(t3$table3 |> mutate(Parameter = gk(Parameter)),
      col.names = c("Parameter","NH Est","NH (SE)","HY Est","HY (SE)","JT Est","JT (SE)"),
      caption = "Table 3. MNL estimates — Nonhyp | Hyp | Joint (with scale).")

mnl_summary <- tibble(
  Statistic = c("Log-likelihood", "Observations", "Scale (mu)", "SE(mu)",
                "Scale-test Chi-sq", "df", "p-value"),
  Nonhyp = c(fmt(t3$ll_nh,1), "1072", "—", "—", "—", "—", "—"),
  Hyp    = c(fmt(t3$ll_hyp,1), "592", "—", "—", "—", "—", "—"),
  Joint  = c(fmt(t3$ll_joint,1), "1664", fmt(t3$mu_hat,3), fmt(t3$se_mu,3),
             fmt(t3$lr_stat,1), as.character(t3$df_test), formatC(t3$p_lr, format="e", digits=1)))
kable(mnl_summary, caption = "Table 3 fit statistics.")
```

::: {.callout-warning}
## Where this differs from the paper
My scale estimate is $\hat\mu = `r fmt(t3$mu_hat,3)`$ (SE $`r fmt(t3$se_mu,3)`$),
against the paper's $0.90$ (SE $0.06$). Both are far enough from 1 to be
significant, and both say the two treatments have different error scales, but the
point estimates are not the same. My scale test is
$\chi^2 = `r fmt(t3$lr_stat,1)`$ on `r t3$df_test` df, while the paper reports
$106.8$ on $4$ df. The $\chi^2$ values are close. The degrees of freedom differ
because of how the joint restriction is counted: I restrict 10 taste parameters
and free 1 scale, which gives $(10+10) - 11 = 9$, whereas the paper's "4" implies
a different count. Either way the conclusion is the same, the pooling is rejected.

One more thing is worth flagging. My log-likelihoods are a bit lower than the
paper's (nonhyp $`r fmt(t3$ll_nh,1)`$ against $-1261.4$; hyp $`r fmt(t3$ll_hyp,1)`$
against $-627.4$). The same offset shows up again in the [IIA test](05-iia.qmd),
and it is the one loose end I could not fully tie off. My best guess is that the
price enters utility a little differently than in the original; I come back to
this in [Discrepancies](08-discrepancies.qmd).
:::
