# Testing IIA with the universal logit {#sec-iia}

```{r}
#| include: false
source("_common.R")
```

## Specification

The universal, or "mother", logit allows the cross-price effects that MNL rules
out:

$$
V_{ij} = \beta_j + \sum_{k=1}^{5} \alpha_{jk}\,P_{ik}.
$$

That is 30 parameters: 5 constants plus a full 5×5 price matrix. MNL is the
special case where $\alpha_{jk}=0$ for $j\neq k$, which is 20 restrictions, so the
IIA likelihood-ratio test has 20 degrees of freedom. A second test takes the
relaxed model and asks whether preferences are equal across treatments, comparing
the joint universal logit with the two separate ones (30 degrees of freedom).

## Estimation code (`R/06_iia_universal_logit.R`)

```{r iia-code, eval=FALSE}
# 25 cross-price dummies: p_{j}_{k} = P_k × 1{alt = j}; zero for NONE.
for (j in steak_alts) for (k in steak_alts)
  df0[[paste0("p_",j,"_",k)]] <- ifelse(df0$alt==j, df0[[paste0("P_",k)]], 0)

f_ul <- as.formula(paste("chosen ~", paste(ul_price_names, collapse=" + "), "| 1"))
ul_nh  <- mlogit(f_ul, make_mdata(filter(df0, treatment=="nonhyp")), reflevel="NONE")
ul_hyp <- mlogit(f_ul, make_mdata(filter(df0, treatment=="hyp")),    reflevel="NONE")
# Joint UL with scale via custom optim (31 params); see script.

chi2 <- function(ll_ul, ll_mnl) 2 * (ll_ul - ll_mnl)   # IIA LR, df = 20
```

## Results

```{r iia-results}
kable(iia$summary_tbl,
      col.names = c("Model","LL(MNL)","LL(UL)","Chi-sq(20)","p-value","Paper Chi-sq","Paper p"),
      caption = "IIA test via universal logit (cf. Lusk & Schroeder 2004, p. 476).")
cat(sprintf("Preference-equality test (relaxed model): Chi-sq(%d) = %.1f, p = %.4f  [paper: Chi-sq(30)=130.7]\n",
            iia$df_eq, iia$chi2_eq, iia$p_eq))
```

::: {.callout-warning}
## Where this differs from the paper
The conclusion is the same as the paper: IIA is rejected in all three samples. But
my test statistics are larger. My $\chi^2(20)$ values are `r fmt(iia$chi2_jt,0)`
(joint), `r fmt(iia$chi2_nh,0)` (nonhyp), and `r fmt(iia$chi2_hyp,0)` (hyp),
against the paper's 66.9, 31.3, and 59.5. The nonhyp case matters most here,
because the paper's 31.3 is only borderline ($p \approx 0.051$) while mine is
clearly significant.

The reason is in the log-likelihoods. My LL levels are consistently lower than the
paper's (for example nonhyp MNL `r fmt(iia$summary_tbl$LL_MNL[2],1)` against
$-1261.4$, and joint UL `r fmt(iia$ll_ul_jt,1)` against $-1908.8$). A roughly
constant gap in the LL level pushes up the IIA $\chi^2$, which is a difference of
two of these offset likelihoods within one sample. It should not affect a test
that differences across samples, and indeed it does not: my preference-equality
test comes out at $\chi^2(30) = `r fmt(iia$chi2_eq,1)`$, almost exactly the
paper's $130.7$. So the problem is in the base MNL/UL specification, most likely
the way price enters utility, rather than in the IIA test itself. More on this in
[Discrepancies](08-discrepancies.qmd).
:::

::: {.callout-note}
## Why IIA matters here
IIA says the relative odds of any two alternatives do not depend on the others.
That is the price of using MNL. Rejecting it, especially in the hypothetical and
pooled samples, is what justifies the richer models in the next chapter: people
are not treating the steaks as independent draws, so a model that lets the error
scales differ across alternatives fits better.
:::
