# Advanced models: HEV and RPL {#sec-table4}

```{r}
#| include: false
source("_common.R")
```

## Specifications

Three models go into Table 4, and I fit two of them.

HEV (heteroscedastic extreme value) is like MNL except that each alternative has
its own Gumbel scale $\sigma_j$, with $\sigma_{\text{NONE}} = 1$ fixed for
identification. Letting the error variances differ by alternative is how it
relaxes IIA. I fit it with a hand-written log-likelihood that integrates the
choice probability over the Gumbel density using trapezoidal quadrature.

RPL (random parameters logit) lets the constants vary across respondents as
normal draws (a mean and a standard deviation each), while the price coefficients
stay fixed so the model converges. I estimate it by simulated maximum likelihood
with 500 draws in `mlogit`.

MNP (multinomial probit) sets the off-diagonal correlations to zero and fixes two
standard deviations to 1. I do not fit it here. It needs the `apollo` package and
a GHK simulator over a five-dimensional normal, which is more than this project
took on, so I describe the intended setup rather than report estimates.

## Estimation code (`R/04_table4_adv.R`)

```{r t4-code, eval=FALSE}
## RPL: ASCs random normal, prices fixed; simulated ML with 500 draws
rpar_spec <- setNames(rep("n", 5), asc_names)
rpl_nh  <- mlogit(f_mnl, mdata_nh,  reflevel="NONE", rpar=rpar_spec, R=500, halton=NA, panel=TRUE)
rpl_hyp <- mlogit(f_mnl, mdata_hyp, reflevel="NONE", rpar=rpar_spec, R=500, halton=NA, panel=TRUE)

## HEV: per-alternative scale sigma_j (sigma_NONE = 1). The probability that j is
## chosen is the integral over the Gumbel density of alternative j of the product
## of the other alternatives' CDFs; evaluated on a trapezoidal grid w in [-8, 45]
## and maximised by optim(BFGS).
opt_hev_nh  <- optim(theta0_nh,  hev_ll, df_s=df_nh,  method="BFGS",
                     control=list(fnscale=-1, reltol=1e-10), hessian=TRUE)
opt_hev_hyp <- optim(theta0_hyp, hev_ll, df_s=df_hyp, method="BFGS",
                     control=list(fnscale=-1, reltol=1e-10), hessian=TRUE)
```

## Results

```{r t4-rpl}
rpl_fit <- tibble(
  Model = c("RPL Nonhypothetical (R=500)", "RPL Hypothetical (R=500)"),
  `Log-likelihood` = c(fmt(t4r$ll_nh, 1), fmt(t4r$ll_hyp, 1)))
kable(rpl_fit, caption = "Table 4a. Random parameters logit fit.")

kable(t4h$table4_hev |> mutate(Parameter = gk(Parameter)),
      col.names = c("Parameter","NH Est","NH (SE)","HY Est","HY (SE)","JT Est","JT (SE)"),
      caption = "Table 4b. HEV estimates — taste (beta, alpha) and scale (sigma) parameters. sigma(NONE) = 1 fixed.")
cat(sprintf("HEV log-likelihood — NH: %.1f | HY: %.1f | Joint: %.1f\n",
            t4h$ll_nh, t4h$ll_hyp, t4h$ll_joint))
```

::: {.callout-warning}
## A convergence problem with HEV
The HEV scale for USDA Choice comes out far too large
($\hat\sigma_{\text{CHO}} \approx `r fmt(t4h$table4_hev$NH_Est[14],0)`$ in the
nonhyp model). That is a sign the variance of this alternative is only weakly
identified given the price design. It feeds through to an inflated HEV WTP for
Choice in the [next chapter](07-wtp-figure.qmd), and it leaves the HEV covariance
matrix close to singular, which is why I run it through `Matrix::nearPD` before
the Poe bootstrap. Better starting values or tighter bounds (I already estimate
$\log\sigma$ rather than $\sigma$) would be the next thing to try. This is the
shakiest number in the whole replication.
:::
