# Advanced models: HEV and RPL {#sec-table4}

```{r}
#| include: false
source("_common.R")
```

## Specifications

- **HEV (heteroscedastic extreme value).** Like MNL but each alternative has its
  own Gumbel scale $\sigma_j$ (with $\sigma_{\text{NONE}} = 1$ fixed for
  identification). It relaxes IIA by allowing alternative-specific error
  variances. I estimate it with a **custom log-likelihood** that integrates the
  choice probability over the Gumbel density by trapezoidal quadrature.
- **RPL (random parameters logit).** ASCs are random-normal across respondents
  (mean + SD); price coefficients are held fixed for convergence. Estimated by
  simulated ML (500 draws) in `mlogit`.
- **MNP (multinomial probit).** Off-diagonal correlations set to 0, two SDs fixed
  to 1. *Not estimated here* — it needs the `apollo` package and a GHK simulator
  over a 5-dimensional normal; I document the intended setup instead.

## Estimation code (`R/04_table4_adv.R`)

```{r t4-code, eval=FALSE}
## RPL: ASCs random normal, prices fixed; simulated ML with 500 draws
rpar_spec <- setNames(rep("n", 5), asc_names)
rpl_nh  <- mlogit(f_mnl, mdata_nh,  reflevel="NONE", rpar=rpar_spec, R=500, halton=NA, panel=TRUE)
rpl_hyp <- mlogit(f_mnl, mdata_hyp, reflevel="NONE", rpar=rpar_spec, R=500, halton=NA, panel=TRUE)

## HEV: per-alternative scale sigma_j (sigma_NONE = 1). The probability that j is
## chosen is the integral over the Gumbel density of alternative j of the product
## of the other alternatives' CDFs; evaluated on a trapezoidal grid w in [-8, 45]
## and maximised by optim(BFGS).
opt_hev_nh  <- optim(theta0_nh,  hev_ll, df_s=df_nh,  method="BFGS",
                     control=list(fnscale=-1, reltol=1e-10), hessian=TRUE)
opt_hev_hyp <- optim(theta0_hyp, hev_ll, df_s=df_hyp, method="BFGS",
                     control=list(fnscale=-1, reltol=1e-10), hessian=TRUE)
```

## Results

```{r t4-rpl}
rpl_fit <- tibble(
  Model = c("RPL Nonhypothetical (R=500)", "RPL Hypothetical (R=500)"),
  `Log-likelihood` = c(fmt(t4r$ll_nh, 1), fmt(t4r$ll_hyp, 1)))
kable(rpl_fit, caption = "Table 4a. Random parameters logit fit.")

kable(t4h$table4_hev |> mutate(Parameter = gk(Parameter)),
      col.names = c("Parameter","NH Est","NH (SE)","HY Est","HY (SE)","JT Est","JT (SE)"),
      caption = "Table 4b. HEV estimates — taste (beta, alpha) and scale (sigma) parameters. sigma(NONE) = 1 fixed.")
cat(sprintf("HEV log-likelihood — NH: %.1f | HY: %.1f | Joint: %.1f\n",
            t4h$ll_nh, t4h$ll_hyp, t4h$ll_joint))
```

::: {.callout-warning}
## Convergence issue — HEV $\sigma_{\text{CHO}}$
The estimated scale for USDA Choice is implausibly large
($\hat\sigma_{\text{CHO}} \approx `r fmt(t4h$table4_hev$NH_Est[14],0)`$ nonhyp).
This signals weak identification of that alternative's variance given the price
design, and it inflates the HEV WTP for Choice in the [next chapter](07-wtp-figure.qmd)
and makes the HEV covariance matrix near-singular (I repair it with
`Matrix::nearPD` before the Poe bootstrap). Better starting values or bounded
reparameterisation ($\log\sigma$ already used) would be the next step. This is the
single largest numerical fragility in my replication.
:::
