# Replication experience and discrepancies {#sec-discrepancies}

```{r}
#| include: false
source("_common.R")
```

This chapter is the part the project asks for directly: what was hard, what I had
to port from LIMDEP to R, and how I read the gaps between my numbers and the
paper's.

## Locating data and documentation

The data file (`harrison.xls`) and the questionnaire (`harrison.doc`) were both
available, but neither was documented. The spreadsheet has a two-row header,
label rows embedded in the middle, and summary rows at the bottom that look like
data until you check them. The cleaning decision that mattered most, dropping
subject 1060, is not obvious unless you go looking for rows with no choice
answers, and that one step is what resolves the 68-versus-67 subject count.

The price design is another example. It lives only in the paper's appendix, not
in the data file. It is the link between a stored choice code and the prices a
respondent actually saw, so a single typo there would corrupt every estimate
downstream. I keyed it in by hand and checked it against the appendix twice.

## Translating LIMDEP → R

`mlogit` has no ready-made command for the scaled joint MNL or for HEV, so I wrote
both as custom log-likelihoods and maximised them with `optim` (BFGS), using the
log-sum-exp trick for stability and trapezoidal quadrature for the HEV Gumbel
integral. In LIMDEP these are canned estimators. Writing them out by hand was
useful for understanding, but it is also where small numerical differences creep
in.

A second point is that the identification choices have to be made explicit in R.
The base alternative (`NONE`), the fixed scale ($\mu_{\text{H}}=1$), and the fixed
HEV scale ($\sigma_{\text{NONE}}=1$) are defaults in LIMDEP but have to be coded
by hand here.

There was also one practical snag worth recording: `dfidx` 0.2.0 together with
`vctrs` triggers a node-stack overflow, so I built the choice data with
`mlogit.data()` instead.

## Interpreting the discrepancies

```{r disc-table, echo=FALSE}
disc <- tibble(
  Item = c("MNL/UL log-likelihood level", "IIA test Chi-sq(20)", "MNL scale mu",
           "Scale-test df", "HEV sigma_CHO", "MNP", "RPL joint model"),
  `My replication` = c(paste0(fmt(t3$ll_nh,0), " (nonhyp MNL)"),
                       paste0(fmt(iia$chi2_nh,0), " (nonhyp)"),
                       fmt(t3$mu_hat,3),
                       paste0(t3$df_test, " (=(10+10)-11)"),
                       paste0("~ ", fmt(t4h$table4_hev$NH_Est[14],0), " (NH)"),
                       "not estimated",
                       "skipped (runtime)"),
  Paper = c("-1261.4", "31.3 (p~0.051)", "0.90", "4", "moderate",
            "estimated (apollo-style)", "estimated"),
  `My reading` = c(
    "Systematic level offset -> likely a price-utility specification difference (main open issue).",
    "Inflated by the LL-level offset; preference-equality test (differenced) matches (127 vs 131).",
    "Same sign & significance; optimiser/start-value sensitivity.",
    "Bookkeeping of the joint restriction differs; conclusion (reject pooling) identical.",
    "Weak identification of Choice's error variance given the price design.",
    "Needs GHK simulator (apollo); out of scope but documented.",
    "Separate-model LLs used as the scale-test benchmark instead."))
kable(disc, caption = "Documented discrepancies and interpretation.")
```

Taking stock, the qualitative side of the paper replicates cleanly. The two
treatment groups are balanced on demographics; the real-payment group chooses
"None" more often; IIA is rejected for the pooled and hypothetical data; and
hypothetical WTP is above real WTP for every steak under every model. The numbers
that do not match are limited to a few things: the log-likelihood level (and the
IIA $\chi^2$ that it inflates), the scale point estimate and the test degrees of
freedom, and the one weakly identified HEV scale. None of these reflects a real
disagreement with the paper; they come from optimiser defaults and from
identification conventions that differ between LIMDEP and R.

::: {.callout-note}
## What I would chase next
The constant log-likelihood gap is the thread I would pull on with more time. The
most likely cause is the price specification. I use five alternative-specific
own-price coefficients $\alpha_j$, but a single generic price coefficient, or a
different way of scaling price, would change the likelihood I can reach while
leaving the qualitative story untouched. Checking this against the paper's Table 3
specification and its footnote 13 would probably close most of the remaining gap.
:::
